---
title: "Srinivasan_Sapflow"
author: "Radha V. Srinivasan & Kendalynn A. Morris"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

# TEMPEST sap flux density overview

```{r Load Packages, include=FALSE}

.packages = c("tidyr", "ggplot2", "dplyr", "lme4",
              "AICcmodavg", "MuMIn", "pbkrtest", "readr",
              "lubridate", "car", "parallel", "data.table",
              "blmeco", "lsmeans", "patchwork", "viridis",
              "forcats", "ggpmisc", "stringr", "e1071", "ggpubr")
#Install CRAN packages (if not already installed)
.inst <- .packages %in% installed.packages()
if(length(.packages[!.inst]) > 0) install.packages(.packages[!.inst])

#Attach packages
sapply(.packages, require, character.only=TRUE)
```

```{r Load Data, message = FALSE, warning = FALSE}
#This will only work once raw data is downloaded and unzipped from ESS-DIVE
#urls to come
#Following unzip, source files create_inventory.R and create_full_timeseries_plus_abiotic.R
#these files calculate sapflux density

#This only needs to be done once
#For some reason, this only works when copy-pasted into the console -R
setwd("/Users/radha/Documents/TEMPEST_Sapflow_Analysis")

final_tmp_data <- readRDS("final_tmp_data.rds")
final_data <- readRDS("final_data.rds")

```

```{r Sapflow Visual, message = FALSE, , warning = FALSE}

#something like this
final_data %>% 
  mutate(Hour = hour(TIMESTAMP)) %>%
  mutate(Date = date(TIMESTAMP)) %>%
  mutate(monthyr = floor_date(TIMESTAMP, unit = "week")) %>%
  filter(Hour <= 19, Hour >= 7) %>%
  filter(`F` <= 0.005, `F` >= 0, 
        PAR >= 700) %>%
  group_by(Plot, Species, Date) %>% 
  summarise(F_avg = mean(`F`, na.rm = TRUE)) -> sf_plot_avg

ggplot(sf_plot_avg) + 
  geom_point(aes (x = Date, y = F_avg, color = Species)) + 
  facet_wrap(~Plot, ncol = 1, scales = "fixed") + 
  scale_color_viridis_d(option = 'D', begin = .25, end = .8) +
  theme_light() +
  labs(y = "Avg Sap Flux Density cm3/cm2/s", x = "Date",
       title = "Sap Flux Density Averaged Daily, 7 am - 7 pm")

```

```{r TEMPEST Events, message = TRUE}
#TEMPEST EVENTS
tempest_events <- bind_rows(
  tibble( Year = 2021, flood_start = "2021-06-12", flood_end = "2021-06-12"), #average date = June 12
  tibble( Year = 2022, flood_start = "2022-06-22", flood_end = "2022-06-22"), #June 22
  tibble( Year = 2023, flood_start = "2023-06-06", flood_end = "2023-06-07"), #June 6, 7
  tibble( Year = 2024, flood_start = "2024-06-11", flood_end = "2024-06-13"), #June 11, 12, 13
  tibble( Year = 2025, flood_start = "2025-06-12", flood_end = "2025-06-12")) #No TEMPEST event; averaging

#growing season from May 15th - October 1st
#I'm sure there's a better way to do this but
growing_szn <- bind_rows(
  tibble(Year = 2021, szn_start = "2021-05-15", szn_end = "2021-10-01",flood_start = "2021-06-12", flood_end = "2021-06-12"),
  tibble(Year = 2022, szn_start = "2022-05-15", szn_end = "2022-10-01", flood_start = "2022-06-22", flood_end = "2022-06-22"),
  tibble(Year = 2023, szn_start = "2023-05-15", szn_end = "2023-10-01", flood_start = "2023-06-06", flood_end = "2023-06-07"),
  tibble(Year = 2024, szn_start = "2024-05-15", szn_end = "2024-10-01", flood_start = "2024-06-11", flood_end = "2024-06-13"),
  tibble(Year = 2025, szn_start = "2025-05-15", szn_end = "2025-10-01", flood_start = "2025-06-12", flood_end = "2025-06-12"))

growing_szn %>%
  mutate(szn_start = ymd(szn_start),
         szn_end = ymd(szn_end),
         flood_start = ymd(flood_start),
         flood_end = ymd(flood_end)) -> season

tempest_events %>%
  mutate(flood_start = ymd(flood_start),
         flood_end = ymd(flood_end)) -> events

#window of data to look at for BACI analysis
window <- days(15)

```


## Two weeks Before/After events

Selecting data from Tulip poplars and excluding the Freshwater plot, based on a preliminary analysis during Radha's SULI project.
```{r Select LITU and SW plot data, message = FALSE, warning = FALSE}
#currently this occurs in the 3 chunks
#may be possible to streamline
#R edits: changing flood window to whole growing season
final_data %>%
  dplyr::select(Year, Species, ID, TIMESTAMP, Plot, `F`,
                soil_vwc_15cm, soil_ec_15cm, swc_sd, ec_sd, PAR) %>%
  mutate(Date = as_date(date(TIMESTAMP)),
         Hour = hour(TIMESTAMP)) %>%
  left_join(season) %>% 
  group_by(Year) %>%
  #mutate(data_start = flood_start - window,
  #       data_end = flood_end + window) %>%
  filter(Date > szn_start & Date < szn_end,
         Hour <= 19, Hour >= 7,
         F <= 0.005, F >= 0,
         (PAR >= 700 | is.na(PAR)),
         Plot != "Freshwater",
         Species == "Tulip Poplar") %>%
  ungroup() -> tulip_salt  

#sapflow, soil ec, soil vwc grouped by day and averaged
#should we change this to hourly averages for ec?
tulip_salt %>%
  group_by(Year) %>%
  mutate(BA = ifelse(Date > flood_end, "After", "Before"),
         Year = as.factor(Year),
         ID = as.factor(ID),
         BA = as.factor(BA),
         Plot = as.factor(Plot)) %>%
  group_by(Date, Plot, ID, Year, BA, flood_start, flood_end, szn_start, szn_end) %>%
  summarise(F_avg = mean(`F`, na.rm = TRUE),
            soil_ec_avg = mean(soil_ec_15cm, na.rm = TRUE), 
            soil_vwc_avg = mean(soil_vwc_15cm, na.rm = TRUE),
            n = n()) %>%
  ungroup() -> tsb_1

tsb_1 %>%
  group_by(Year) %>%
  mutate(Day = ifelse(BA == "Before",
                      - as.numeric(flood_start - Date),
                      as.numeric(Date - flood_end))) %>%
  mutate(BA = as.character(BA)) %>%
  mutate(BA = ifelse(Year == 2021, "Before", BA),
         BA = ifelse(Year == 2025, "Before", BA)) -> tsb_2

ggplot(tsb_2, aes(Day, F_avg, color = ID, shape = BA)) +
  geom_point(size = 2.5) + facet_grid(Plot~Year, scales = "free")
```

```{r Normality, message = FALSE}
#here we check if our response variable (daily averaged midday sap flux density per tree), is normally distributed
current_skewness <- skewness(tsb_2$F_avg)
print(paste("Current Skewness:", current_skewness))

# Logarithmic Transformation
log_transformed_response <- log(tsb_2$F_avg + 1)

# Square Root Transformation
sqrt_transformed_response <- sqrt(tsb_2$F_avg)

# Inverse Transformation
inv_transformed_response <- 1 / tsb_2$F_avg

log_skewness <- skewness(log_transformed_response)
sqrt_skewness <- skewness(sqrt_transformed_response)
inv_skewness <- skewness(inv_transformed_response)

print(paste("Log Transformation Skewness:", log_skewness))
print(paste("Square Root Transformation Skewness:", sqrt_skewness))
print(paste("Inverse Transformation Skewness:", inv_skewness))

ggplot(tsb_2, aes(F_avg)) + geom_histogram(fill = '#31688EFF') +
  facet_grid(Plot~.) +
  labs(y = NULL, x = "Average Sapflow", title = "Untransformed Data")
  

ggplot(tsb_2, aes(log(F_avg + 1))) + geom_histogram(fill = '#31688EFF') +
  facet_grid(Plot~.) + 
  labs(y = NULL, x = "Avergae Sapflow", title = "Log \n Transformed Data")

#Log transformation is ever so slightly less skewed but looks virtually identical, just keeping it untransformed for now
```

```{r}
#quick t-test for ec differences: Control group has significantly higher ec than saltwater pre-flood
final_tmp_data %>%
  filter(Plot != "Freshwater", 
         Year == "2021") -> temp2
 
 t.test(soil_ec_15cm ~ Plot, data = temp2)
```


# Progressive Analysis



We follow the workflow developed by Pardini et al 2018 [pardini_link](https://doi.org/10.1111/rec.12678)  and start with the following linear mixed model:

sqrt(F_avg) \~ BA + Plot + BA\*Plot + (1\|ID) + (1\|Year) + (1\|ID:Year)

In which 'Plot' is the control/impact treatment effect (the Control and Saltwater plots for TEMPEST), 'BA' is the before and after treatment effect (pre/post flooding events), and tree ID and Year are random effects.

\*Note that although, flooding intensity increases each year, we ignore this effect for now.

First, test for the statistical significance of the final term (1\|ID:Year), the interaction of the ID and Year random effects, using an AIC comparison.

```{r Pardini et al 2018, warning = FALSE}
Cand.set <- list()

Cand.set[[1]] <- lmer(sqrt(F_avg) ~ BA + Plot  + BA*Plot +
                         (1|ID) + (1|Year),
                       data = tsb_2)
model.noint <- Cand.set[[1]]

Cand.set[[2]] <- lmer(sqrt(F_avg) ~ BA + Plot + BA*Plot +
                         (1|ID) + (1|Year) +
                         (1|ID:Year),
                       data = tsb_2)
model.int <- Cand.set[[2]]

AIC.res.table <- aictab(cand.set = list(Cand.set[[1]], Cand.set[[2]]), 
                        modnames = paste0("Cand.set_", c(1,2)), 
                        second.ord = TRUE)
AIC.res.table
```

Choose the model with the lowest AICc value (i.e. largest negative value); this suggests that the second model, which includes the random effects interaction term, is a better fit for our data.


# Create and test BACI models


We now test the impact of the BACI term in our existing model by looking at the explanatory power of the model with and without the interaction between BA and CI (plot): 
```{r message = FALSE, warning = FALSE}
model.int.og <- lmer(sqrt(F_avg) ~ BA*Plot +
                        (1|ID) + (1|Year) + (1|ID:Year),
                      data = tsb_2)

model.noint.og <- lmer(sqrt(F_avg) ~ BA + Plot +
                          (1|ID) + (1|Year) + (1|ID:Year),
                        data = tsb_2)

og.refdist <- PBrefdist(largeModel = model.int.og, 
                        smallModel = model.noint.og, 
                        nsim = 100)

og.compar <- PBmodcomp(largeModel = model.int.og, 
                       smallModel = model.noint.og,
                       ref = og.refdist)

og.compar
Anova(model.int.og, type = "III")
```

Neither the likelihood ratio test nor the parametric bootstrap test show significance for the BACI interaction term.
# R edits: with unscaled 2025 data included, it looks like LRT and PB do both show significance for the BACI term this time around!

```{r}
#Sensor ID is proving difficult to add in, but could test difference with:

#ec.test <- lmer(soil_ec_avg ~ BA*Plot*Year + 
                # (1:Sensor_ID),
               # data = tsb_2)
```


As a proxy for the increase in flooding intensity with each year, we include soil electrical conductivity into the model. Note that soil EC is an indicator of salt accumulation in soil from brackish water application: 

```{r}
#BACI with just soil EC to account for pre-existing differences: 
model.int.ec <- lmer(soil_ec_avg ~ BA*Plot +
                      (1|Year),
                      data = tsb_2)

model.noint.ec <- lmer(soil_ec_avg ~ BA + Plot +
                      (1|Year),
                      data = tsb_2)

ec.refdist <- PBrefdist(largeModel = model.int.ec, 
                        smallModel = model.noint.ec, 
                        nsim = 100)

ec.compar <- PBmodcomp(largeModel = model.int.ec, 
                       smallModel = model.noint.ec,
                       ref = ec.refdist)
ec.compar
Anova(model.int.ec, type = "III")

```

```{r BACI plus EC, message = FALSE, warning = FALSE}
model.int.ec <- lmer(F_avg ~ BA*Plot + soil_ec_avg +
                      (1|ID) + (1|Year) + (1|ID:Year),
                      data = tsb_2)

model.noint.ec <- lmer(F_avg ~ BA + Plot + soil_ec_avg +
                        (1|ID) + (1|Year) + (1|ID:Year),
                        data = tsb_2)

ec.refdist <- PBrefdist(largeModel = model.int.ec, 
                        smallModel = model.noint.ec, 
                        nsim = 100)

ec.compar <- PBmodcomp(largeModel = model.int.ec, 
                       smallModel = model.noint.ec,
                       ref = ec.refdist)
ec.compar
Anova(model.int.ec, type = "III")
```

We now see that the BACI interaction term and soil ec fixed effect have strong explanatory power via the likelihood ratio and parametric bootstrap 

#R edits: still statistically signficant, slightly more so than orginal BACI without soil ec


# Plots


AIC comparison to visually observe differences in how well each model fits the data: 

```{r AIC, message = FALSE, warning = FALSE}

model.noint.random <- lmer(sqrt(F_avg) ~ BA*Plot + (1|ID) + (1|Year),
                           data = tsb_2)

model.random.int <- lmer(sqrt(F_avg) ~ BA*Plot + (1|ID) + (1|Year) +
                           (1|ID:Year),
                         data = tsb_2)

model.ec.only <- lmer(sqrt(F_avg) ~ soil_ec_avg +
                        (1|ID) + (1|Year) + (1|ID:Year),
                         data = tsb_2)


AIC(model.noint.random) -> AIC_initial
AIC(model.random.int) -> AIC_randominteraction
AIC(model.int.ec) -> AIC_int_ec #BACI interaction plus SEC
AIC(model.noint.ec) -> AIC_noint_ec #BA + CI plus SEC
AIC(model.ec.only) -> AIC_ec_only


variable <- c("Model A", "Model B", "Model C", "Model D", "Model E")
AIC <- c(AIC_initial, AIC_randominteraction,
         AIC_int_ec, AIC_noint_ec, AIC_ec_only)
candidate <- c("without interaction in random effects",
               "with interaction in random effects", 
               "with Soil EC",
               "no BACI interaction with Soil EC",
               "just Soil EC")

AIC <- data.frame(variable, AIC, candidate)

AIC %>%
  mutate(`Model Specification` = candidate) %>%
  ggplot(aes(variable, AIC+5000, fill = `Model Specification`)) +
  geom_col() +
  geom_hline(yintercept = -2820, linetype = "dashed") +
  expand_limits(y = c(0, -3000)) +
  scale_y_continuous(expand = c(0, 0)) + # Starts x axis at true zero
  scale_fill_viridis_d() +
  labs( title = 'AIC Model Comparison', x = 'Variable', y = 'AIC Value + 5000') +
  guides(fill = FALSE) + 
  theme_light() 
```

```{r Sapflow vs EC, message = FALSE, warning = FALSE}
tsb_2_summary = tsb_2 %>%
  filter(soil_ec_avg < 1250) %>%
  mutate(BA = case_when(Year == 2021 ~ "Before", .default = BA),
         BA = factor(BA, levels = c("Before", "After")),) %>%
  group_by(Year,BA,Plot) %>%
  summarise(mean_annual_F = mean(F_avg),
            sd_annual_F = sd(F_avg),
            mean_annual_EC = mean(soil_ec_avg),
            sd_annual_EC = sd(soil_ec_avg),
            n = n())
tsb_2 %>%
  mutate(BA = factor(BA, levels = c("Before", "After"))) -> tsb_2

tsb_2_summary %>%
  ggplot() +
  stat_smooth(aes(x = soil_ec_avg, y = F_avg), method = 'lm',
              color = 'grey', size = 0.8,
              data = tsb_2 %>% filter(soil_ec_avg < 1200)) +
  geom_point(aes(mean_annual_EC, mean_annual_F,
                 color = as.factor(Year)), size = 3) +
  geom_errorbar(aes(x = mean_annual_EC,
                    ymin = mean_annual_F - sd_annual_F,
                    ymax = mean_annual_F + sd_annual_F,
                    color = Year), size = 1) +
  geom_errorbarh(aes(xmin = mean_annual_EC - sd_annual_EC,
                     xmax = mean_annual_EC + sd_annual_EC,
                     y = mean_annual_F,
                     color = Year), size = 1) +
  scale_color_viridis_d(begin = 0.9, end = 0.2) +
  facet_wrap(Plot ~ BA, scales = "free_x") + theme_bw() +
  geom_point(aes(x = soil_ec_avg, y = F_avg, color = as.factor(Year)),
             data = tsb_2 %>% filter(soil_ec_avg < 1200), alpha = 0.2) +
  labs(y = expression ("Midday Sap Flux Density, cm"^3*" cm"^-2*" s"^-1),
       x = expression(paste("Mean Soil EC, ", mu, "S cm"^-1)),
       color = " ") +
  theme(legend.position="bottom", element_text(size = 14))
```

```{r VWC vs EC, message = FALSE, warning = FALSE}

tsb_2_summary2 = tsb_2 %>%
  filter(soil_ec_avg < 1250) %>%
  mutate(BA = case_when(Year == 2021 ~ "Before", .default = BA),
         BA = factor(BA, levels = c("Before", "After"),ordered = TRUE)) %>%
  group_by(Year,BA,Plot) %>%
  summarise(mean_annual_vwc = mean(soil_vwc_avg),
            sd_annual_vwc = sd(soil_vwc_avg),
            mean_annual_EC = mean(soil_ec_avg),
            sd_annual_EC = sd(soil_ec_avg),
            n = n())

tsb_2_summary2 %>%
  ggplot() +
  # stat_smooth(aes(x = soil_ec_avg, y = soil_vwc_avg, group = BA),
  #             method = 'lm', se = FALSE,
  #             color = 'grey', size = 0.8,
  #             data = tsb_2 %>% filter(soil_ec_avg < 1200)) +
  geom_point(aes(mean_annual_EC, mean_annual_vwc,
                 color = as.factor(Year)), size = 3) +
  geom_errorbar(aes(x = mean_annual_EC,
                    ymin = mean_annual_vwc - sd_annual_vwc,
                    ymax = mean_annual_vwc + sd_annual_vwc,
                    color = Year), size = 1) +
  geom_errorbarh(aes(xmin = mean_annual_EC - sd_annual_EC,
                     xmax = mean_annual_EC + sd_annual_EC,
                     y = mean_annual_vwc,
                     color = Year), size = 1) +
  geom_point(aes(x = soil_ec_avg, y = soil_vwc_avg, color = as.factor(Year)),
             data = tsb_2 %>% filter(soil_ec_avg < 1200), alpha = 0.2) +
  scale_color_viridis_d(option = 'D', begin = 0.9, end = 0.2) +
  facet_wrap(Plot ~ BA, scales = "free_x") + theme_bw() +
  labs(y = expression ("Mean Soil Volumetric Water Content cm" ^3* " cm"^-3),
       x = expression(paste("Mean Soil EC, ", mu, "S cm"^-1)),
       color = " ") +
  theme(legend.position="bottom", element_text(size = 14))
```
